Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community

### 2020-10-24 manuscript.rmd
1. formatted equations in mortality analyses 2. presented stats in mortality results 2. as well as ERD vs. iso results

\begin{equation}
\begin{split}
B_{s} = `r bcf[1]` `r bcf[2]`\textrm{WSG}_{s} + `r bcf[3]`\textrm{WSG}_{s}^2 
& `r bcf[4]`\textrm{WSG}_{s}\textrm{LMA}_{s} + `r bcf[5]`\textrm{LMA}_{s} `r bcf[6]`\textrm{LMA}_{s}^2 + \epsilon_{s,b}
(\#eq:b-mod)
\end{split}
\end{equation}
where $\epsilon_{s,a}$ and $\epsilon_{s,b}$ are error terms.

@@

We calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$30 m) and average abundance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; $n$ = `r length(erd.mrate.sp)`). For each of the seven census intervals $t$ in the 35-yr record (1981-82, 1985, 1990, 2000, 2005, 2010, 2015, 2015), $\textrm{Mort}_{t}$ (% yr^-1^) for species $s$ for two successive censuses, $c_{1}$ and $c_{2}$, was calculated as $\textrm{Mort}_{s,t} = \frac{D_{s,c_{1}}}{N_{s,c_{2}}}\times\frac{100}{\textrm{d}}$, where $N_{s,c_{1}}$ and $D_{s,c_{2}}$ are the total number of trees of a species $s$ present in $c_{1}$ and dead in $c_{2}$, respectively, in the 50 ha plot and $d$ is the duration based on mean dates of $c_{1}$ and $c_{2}$.

@@

\
Fig. \@ref(fig:hydro-calib) here.

@@

\
Fig. \@ref(fig:swp-droughts) here.

@@

ERD estimated from the chosen model with $K_\textrm{leaf}$ and a multiplicative effect of VPD as explanatory variables (Eq. \@ref(eq:M2)) was strongly related to the measured isotopic signal ($\mathrm{\delta}$^2^H), that is, with a high goodness-of-fit ($R$^2^ = 0.65; $p$ = 0.027, $n$ = `r length(unique(erd.iso$sp))`; Fig. \@ref(fig:erd-iso)). Two of the alternative models tested--with $K_\textrm{leaf}$ and LAI, with or without an effect of VPD--had a strong relationship with $\mathrm{\delta}$^2^H as well, but one that was based on a smaller set of species ($n$ = 4) as fewer species passed the criteria of sufficient species-level goodness-of-fit; <20 vs. `r length(erd.sp)` species for the chosen model (Fig. S\@ref(fig:other-erd-mods)). For the rest of the analyses we therefore used ERD estimates from the less parimonious, chosen model (Eq. \@ref(eq:M-chosen)).

\
Fig. \@ref(fig:erd-iso) here.

\   In this forest, modeled effective rooting depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from `r min(df.erd.to.plot$depth)` m to `r max(df.erd.to.plot$depth)` m (Fig. \@ref(fig:sp-erd)). Notably, two species, *L. seemannii* and *T. tuberculata*, that @Meinzer:1999 found to have xylem sap $\mathrm{\delta}$^2^H values between soil water and groundwater, suggesting that these species sourced most of the water from depths >1 m, were also identified by our model with ERD >1 m (2.9 m for both species; Fig. \@ref(fig:erd-iso)).

@@

\
Fig. \@ref(fig:erd-hyd) here.

@@

Species with deeper ERD had significantly lower mortality rates than deeper ERD species in the census intervals 1985-1990, 2000-2005 and 2005-2010. ($n$ = `r nrow(mrate.depth.mean)`, Fig. S\@ref(fig:erd-mort-census)). These relationships were stronger among evergreen species ( $n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`; Fig. \@ref(fig:erd-mort-census-evg)). Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme soil droughts distinctly occurred in the wet season, effectively extending the dry season (Fig. \@ref(fig:swp-droughts)). Soil water availability was not modelled for 1985-1990. Analysis of a species' time spent under significant soil drought (proportion of days $\Psi_{\textrm{soil},z=\textrm{ERD}}$ < $\Psi$~crit~) showed that species with shallower ERD spent greater proportion of days below critical hydraulic thresholds (Fig. \@ref(fig:erd-time); S\@ref(fig:erd-time-evg)). Species with ERD deeper than 1 m did not cross species $\Psi$~crit~ for the entire 25-yr period for which soil water potential $\Psi_{\textrm{soil},z}$ was modeled (1990-2015). Soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`;  median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi$~crit~) of the most sensitive tree species (-0.28 MPa).

\
Fig. \@ref(fig:erd-mort-census-evg) here.
Fig. \@ref(fig:erd-time) here.

@@

The trade-offs that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests. The hydraulic efficiency-safety trade-off between high values of maximum stem conductivity (high $K_\textrm{max,stem}$) and cavitation-resistance (low $\Psi$~50,stem~ or $\Psi$~88,stem~) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016]. We report such a trade-off ($n$ = 26 species) at our study site stem level (). We also found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown). WSG alone explained `r round(a.wsg.r2*100,0)` % of variation in parameter $A$ ($A_{s} = `r a.wsg.cf[1]` `r a.wsg.cf[2]`B_{s} + `r a.wsg.cf[3]`B^2_{s} + \epsilon_{s}$; $p$ `r  a.wsg.p`) and `r round(b.wsg.r2*100,0)` % of variation in $B$ ($B_{s} = `r b.wsg.cf[1]` `r b.wsg.cf[2]`B_{s} + `r b.wsg.cf[3]`B^2_{s} + \epsilon_{s}$; $p$ `r  b.wsg.p`). Parameter $B$ alone explained `r round(a.b.r2*100,0)` % of variation in $A$ ($p$ `r  a.b.p`), while LMA alone was weakly related to $A$ and was not related to $B$ (not shown). The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.

@@

\begin{equation}
G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*}\textrm{VPD}_{i}^{*}\textrm{LAI}_{doy|i}^{*}) + \epsilon_{t|z}
(\#eq:M6)
\end{equation}
where n~t~ be the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is a normally distributed model error term.

>>>>

\begin{equation}
\begin{split}
B_{s} = `r bcf[1]` `r bcf[2]`\textrm{WSG}_{s} + `r bcf[3]`\textrm{WSG}_{s}^2 
&`r bcf[4]`\textrm{WSG}_{s}\textrm{LMA}_{s} + `r bcf[5]`\textrm{LMA}_{s} `r bcf[6]`\textrm{LMA}_{s}^2 + \epsilon_{s,b}
(\#eq:b-mod)
\end{split}
\end{equation}
where $\epsilon_{s,a}$ and $\epsilon_{s,b}$ are error terms.

@@

To test whether ERD plays a role in alleviating mortality risk, ee calculated mortality rates for large trees ($\ge$ 10 cm dbh) of species with ERD estimates (all canopy species with maximum height $\ge$30 m) and average abundance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; $n$ = `r length(erd.mrate.sp)`). For each of the seven census intervals $t$ in the 35-yr record (1981-82, 1985, 1990, 2000, 2005, 2010, 2015, 2015), $\textrm{Mort}_{t}$ (% yr^-1^) for species $s$ for two successive censuses, $c_{1}$ and $c_{2}$, was calculated as $\textrm{Mort}_{s,t} = \frac{D_{s,c_{1}}}{N_{s,c_{2}}}\times\frac{100}{\textrm{d}}$, where $N_{s,c_{1}}$ and $D_{s,c_{2}}$ are the total number of trees of a species $s$ present in $c_{1}$ and dead in $c_{2}$, respectively, in the 50 ha plot and $d$ is the duration based on mean dates of $c_{1}$ and $c_{2}$.

@@

Fig. \@ref(fig:hydro-calib) here.

@@

Fig. \@ref(fig:swp-droughts) here.

@@

The best model (with a mulicative effect of VPD and not LAI; Eq. \@ref(eq:M-chosen)) explained a large fraction of the variance in the xylem sap isotopic signature ($\mathrm{\delta}$^2^H); $R$^2^ = `r lm.erd.iso.r2`; $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`; Fig. \@ref(fig:erd-iso)). Two of the alternative models tested with LAI had a strong relationship with $\mathrm{\delta}$^2^H as well, but one that was based on a smaller set of species ($n$ = 4) as fewer species passed the criteria of sufficient species-level goodness-of-fit with these models; <20 speceis vs. `r length(erd.sp)` species for the best model (Fig. S\@ref(fig:other-erd-mods)). For the rest of the analyses we therefore used ERD estimates from the less parimonious, best model (with VPD and not LAI, Eq. \@ref(eq:M-chosen)).

Fig. \@ref(fig:erd-iso) here.

\
\   In this forest, modeled effective rooting depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from `r min(df.erd.to.plot$depth)` m to `r max(df.erd.to.plot$depth)` m (Fig. \@ref(fig:sp-erd)). Notably, two species, *L. seemannii* and *T. tuberculata*, that @Meinzer:1999 found to have xylem sap $\mathrm{\delta}$^2^H values between soil water and groundwater, suggesting that these species sourced most of the water from depths >1 m, were also identified by our model with ERD >1 m (2.9 m for both species; Fig. \@ref(fig:erd-iso)).

@@

Fig. \@ref(fig:erd-hyd) here.

@@

Amongst the evergreen species ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 2000-2005 and 2005-2010, respectively, such that species with deeper ERD had significantly lower mortality rates than deeper ERD species ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; $n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`; Fig. \@ref(fig:erd-mort-census-evg)). When deciduous species were included, ERD explained lower but significant proportion of variation in mortality in those intervals ($R$^2^ = `r m.r2[1]`, `r m.r2[2]` and `r m.r2[3]`% and $p$ = `r m.p.90`, `r m.p.05` and `r m.p.10`, respectively; $n$ = `r nrow(mrate.depth.mean)`; Fig. \@ref(fig:erd-mort-census)). Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme soil droughts distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig. \@ref(fig:swp-droughts)). Soil water availability was not modelled for 1985-1990. Analysis of a species' time spent under significant soil drought (proportion of days $\Psi_{\textrm{soil},z=\textrm{ERD}}$ < $\Psi$~crit~) showed that species with shallower ERD spent greater proportion of days below critical hydraulic thresholds (Fig. \@ref(fig:erd-time); S\@ref(fig:erd-time-evg)). Species with ERD deeper than 1 m did not cross species $\Psi$~crit~ for the entire 25-yr period for which soil water potential $\Psi_{\textrm{soil},z}$ was modeled (1990-2015). Soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`;  median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi$~crit~) of the most sensitive tree species (-0.28 MPa).

Fig. \@ref(fig:erd-mort-census-evg) here.
Fig. \@ref(fig:erd-time) here.

@@

The trade-offs that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests. The hydraulic efficiency-safety trade-off between high values of maximum stem conductivity (high $K_\textrm{max,stem}$) and cavitation-resistance (low $\Psi$~50,stem~ or $\Psi$~88,stem~) is hypothesized although community-wide examples are rare [@Hacke:2006; @Gleason.2016]. We report such a trade-off ($n$ = 26 species) at our study site stem level. We also found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown). WSG alone explained `r round(b.wsg.r2*100,0)`% of variation in $B$ ($p$ `r  b.wsg.p`). The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.

@@

\begin{equation}
G_{t|z} = a_{0} + a_{1}(\frac{1}{n_{t}}\sum_{i=1}^{n_{t}} K_{\textrm{leaf},i|z}^{*}\textrm{VPD}_{i}^{*}\textrm{LAI}_{doy|i}^{*}) + \epsilon_{t|z}
(\#eq:M6)
\end{equation}
where n~t~ are the total number of days in census interval $t$, the superscript * indicates that the variable has been standardized between 0 and 1, and $\epsilon$ is the model error term.
